Versioned data: why it is needed and how it can be achieved (easily and cheaply)

### 2017-09-25 ms.tex
more clean-up

At current there are two classes of data users: those that  interact with the data via point and click downloading and those that use code-based interaction.  The system we describe allows for both types (Figure \ref{fig:technology_stack} and Table \ref{tab:user_requirements}).  \texttt{GitHub} has a feature called ``releases'' in which specific commits can be turned into a version for distribution (with associated version number).  Those releases can be downloaded directly by users via point-and-click, or accessed via the \texttt{GitHub} \textsc{api}.
% I have no idea what this next sentence means.
Releases can also be embedded into any kind of website that allows the data curator to facilitate discovery by more users and to provide context.  Releases can also be linked to digital object identifiers and code-based users via for example an \textsc{R} package.

@@

For example, \texttt{datastorr} was used to build the package \texttt{baad.data} (\smurl{github.com/traitecoevo/baad.data}), which is an interface to the Biomass and Allometry Database (\textsc{BAAD}) \citep{Falster-2015} stored at \smurl{github.com/dfalster/baad}. \texttt{baad.data} consists of only a few simple functions, that were automatically generated (along with the associated help files) with \texttt{datastorr}. And importantly \texttt{baad.data} contains no actual data so is very quick to install and takes up virtually no space on the user's hard-drive.
% Dan Noble - Next 5 lines: I found this part to be a bit difficult to grasp as you discuss how updated versions can "download the latest version”, but then in the same sentence discuss how this works seamlessly “offline”. You may want to discuss this in greater depth as I don’t see how this is possible.
When the function \texttt{baad.data::baad\_data()} is called, the system will download the latest version; if there have not been any changes since the last time the function was called, it will simply retrieve the data from memory, making the system very efficient (and also allowing the system to work seamlessly if the computer is offline). If a researcher requires an older version of \textsc{BAAD}, specifying the version number of the data as an argument, e.g. \texttt{baad.data::baad\_data("1.0.1")} will retrieve it, without overwriting any other dataset versions they may have previously downloaded. This enables users to reproduce an analysis using a specific version of the data, even while new data is posted.

@@

Then as the dataset grows over time, researchers update the \texttt{git} repo and create a \texttt{GitHub} release with a new version number (preferably
using \href{http://semver.org/}{semantic versioning}). All the versions
will be available to the user.

% Something not mentioned; that the versions are cached between runs
% so that the data are accessed by name/version (rather than by
% homebrewing a local cache and manually doing a bunch of file name
% mangling) and that all accesses after the first are very fast

We have created \texttt{datastorr} to make \textsc{vdd} easy to implement and manage for \texttt{R} users. And as we stated above, this has already been used to release a variety of datasets (Table \ref{tab:examples}).
% Reword next senetnce
However, we can imagine alternative methods for interacting with \texttt{GitHub} and building templates and adopting a \textsc{vdd} system does not critically depend on using \texttt{datastorr} or \texttt{GitHub}. We would love to see implementations in \texttt{Python} or other languages and using other cloud platforms for data distribution.

@@

% WKC: hmmmm not sure we've nailed this yet, but see
% https://github.com/traitecoevo/data_versioning/issues/10#issuecomment-317059819

% subsection only temporary as not allowed by journal guidelines.
\subsection{Future directions}

Technology for data management and distribution is moving forward rapidly.  The workflow we describe above works well with current tools and should work even more quickly with its future replacement.  Here we highlight a few rapidly developing future tools that researchers should be aware of.

Because \texttt{git} was built for tracking changes in code, there are aspects of how it behaves that are not natural to apply to  data \cite{Perkel-2016}.  For example, because it tracks changes line-by-line rather than cell-by-cell, there are inefficiencies in how the software stores the changes.
% Dan Noble: I think this is a pragmatic view, but I would have a tendency not to discuss how the tools you can use for a VDD approach are likely to be "replaced in the near future”. Of course, things change, but this will not entice the people you want the paper to appeal to to adopt this sort of approach. From my experience, some of the resistance to adopting this stuff comes from programs or repositories being “replaced” at such a high rate. They just don’t feel like its worth doing.
There is rapid current work on new implementations which will be faster computationally, smaller in file size, flexible with respect to data structures, and thus allow for larger datasets to be versioned efficiently\citep{Fli, Dat}.
% Add CKAN here?
That said...
% Add CKAN and Git-LFS, and say something useful about the future

@@

>>>>

At current there are two classes of data users: those that  interact with the data via point and click downloading and those that use code-based interaction.  The system we describe allows for both types (Figure \ref{fig:technology_stack} and Table \ref{tab:user_requirements}).  \texttt{GitHub} has a feature called ``releases'' in which specific commits can be turned into a version for distribution (with associated version number).  Those releases can be downloaded directly by users via point-and-click, or accessed programmatically via the \texttt{GitHub} \textsc{api}.  This system also allows the curator to do internal work between releases and only \emph{release} the data to users when the revision represents a clear improvement on the previous release.

% I have no idea what this next sentence means.  WKC: cut
%Releases can also be embedded into any kind of website that allows the data curator to facilitate discovery by more users and to provide context.  Releases can also be linked to digital object identifiers and code-based users via for example an \textsc{R} package.

@@

For example, \texttt{datastorr} was used to build the package \texttt{baad.data} (\smurl{github.com/traitecoevo/baad.data}), which is an interface to the Biomass and Allometry Database (\textsc{BAAD}) \citep{Falster-2015} stored at \smurl{github.com/dfalster/baad}. \texttt{baad.data} consists of only a few simple functions, that were automatically generated (along with the associated help files) with \texttt{datastorr}. And importantly \texttt{baad.data} contains no actual data so is very quick to install and takes up virtually no space on the user's hard-drive.
% Dan Noble - Next 5 lines: I found this part to be a bit difficult to grasp as you discuss how updated versions can "download the latest version”, but then in the same sentence discuss how this works seamlessly “offline”. You may want to discuss this in greater depth as I don’t see how this is possible.
When the function \texttt{baad.data::baad\_data()} is called, the system will download the latest version; if there have not been any changes since the last time the function was called, it will simply retrieve the data from memory, making the system very efficient (and also allowing the system to work seamlessly if the computer is offline). If a researcher requires an older version of \textsc{BAAD}, specifying the version number of the data as an argument, e.g. \texttt{baad.data::baad\_data("1.0.1")} will retrieve it, without overwriting any other dataset versions they may have previously downloaded. This enables users to reproduce an analysis using a specific version of the data, even while new data is posted.  Users may also have several versions of the same database on their computer and unambiguously access the different versions with one simple function call. The dataset versions are cached between runs such that the data are accessed by name and version, so that after the first download of each version access to the data should be very fast.

% Something not mentioned; that the versions are cached between runs
% so that the data are accessed by name/version (rather than by
% homebrewing a local cache and manually doing a bunch of file name
% mangling) and that all accesses after the first are very fast
% WKC: I had a go at this--see above

@@

Then as the dataset grows over time, researchers update the \texttt{git} repo and create a \texttt{GitHub} release with a new version number (preferably
using \href{http://semver.org/}{semantic versioning}). All the releases are simultaneously
available to the user by both point-and-click and programmatically.

We have created \texttt{datastorr} to make \textsc{vdd} easy to implement and manage for \texttt{R} users. And as we stated above, this has already been used to release a variety of datasets (Table \ref{tab:examples}).
% Reword next senetnce
However, we can imagine alternative methods for interacting with \texttt{GitHub} and building templates and adopting a \textsc{vdd} system.  The key is not the specific technology, but rather the concept of creating, maintaining, and distributing versioned data.

@@

% WKC: hmmmm not sure we've nailed this yet, see
% https://github.com/traitecoevo/data_versioning/issues/10#issuecomment-317059819

% subsection only temporary as not allowed by journal guidelines.
% trick is to say something about how things are getting better rapidly, but not scare people away from doing this now.  
\subsection{Future directions}

Technology for data management and distribution is moving forward rapidly.  The workflow we describe above works well with current tools and should work even more quickly with its future replacement.  Here, we highlight a few rapidly developing future tools that researchers should be aware of.

Because \texttt{git} was built for tracking changes in code, there are aspects of how it behaves that are not natural to apply to  data \cite{Perkel-2016}.  For example, because it tracks changes line-by-line rather than cell-by-cell, there are inefficiencies in how the software stores the changes.
There is rapid current work on new implementations which will be faster computationally, smaller in file size, flexible with respect to data structures, and thus allow for larger datasets to be versioned efficiently\citep{Fli, Dat}.
% Add CKAN, rdrop2 and Git-LFS, conceptually at least, and say something useful about the future

@@

% Dan Noble: I think this is a pragmatic view, but I would have a tendency not to discuss how the tools you can use for a VDD approach are likely to be "replaced in the near future”. Of course, things change, but this will not entice the people you want the paper to appeal to to adopt this sort of approach. From my experience, some of the resistance to adopting this stuff comes from programs or repositories being “replaced” at such a high rate. They just don’t feel like its worth doing.
